### 2021-11-20 draft.tex
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@b110a82e6c34a0498ade3712a5140fd3878765b4 🚀
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@78c1758c6fc8e8d2ce5414417989df6bd54ac5f3 🚀

Within the last several hundred years, human activity has rapidly
changed Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrestial surface and its oceans, and both agricultural and urban
development has rapidly reshaped the cover of Earth's surface. These the
bulk of this change has occurred within the last several hundred years,
a geological instant, potentially inducing shocks to ecosystems that
could threated their integrity (\textbf{Scheffer?}). As a result
understanding and predicting how ecosystems will change in the future,
\emph{ecological forecasting}, and making making descisions based on
these predictions mitigating the consequences of this change, on
ecosystems has emerged as an imperative for ecology and environmental
science {[}Dietze (2017);{]}.

@@

However, robust forecasting of ecological processes will change in the
future is, to say the least, quite difficult (Beckage \emph{et al.}
2011; Petchey \emph{et al.} 2015). This difficultly is compounded by a
few factors, the first being that sampling ecosystems is not easy.
Ecological data is often biased, and noisey, spatially and temporally
sparse. As a result \emph{ecosystem monitoring} (Makiola \emph{et al.}
2020) has emerged as an imperative. Developing a system for ecological
observation, which is able to coordinate across locations.
(\textbf{AndyUrbanBiomonitoring?} paper).

The second major challenge in forecasting ecosystems is that the
underlying dynamics of most ecological processes are fundementally
unknown (and unknowable) and instead must be inferred.

Much of the history of quantitatively modeling ecosystems have been done
in the language of dynamical systems, describing how the value of an
observable state of the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time. It turns
out to be much more effective to, rather than attempt to directly model
\(\vec{x}(t)\) itself, to instead describe how \(\vec{x}\) changes from
one timestep to the next, yielding models in the form of differential
equations in continuous-time settings--\(\frac{dx}{dt} = f(x)\)-- or
difference equations in discrete-time
settings--\(x_t = f(x_{t-1})\)--where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Chen \emph{et al.} 2019).

@@

However, we run into many problems when aiming to apply this type of
model to empirical data in ecology.

The initial success of ODE models can be traced back to the larger
program of ontological reductionism, which became the de facto apporoach
model physical sciences after its early success in physics, which, by
the time ecology was becoming a quantitative science (sometime in the
20th century, depending on who you ask), became the foundation for early
quantitative models in ecology.

But ecosystems are perhaps the quintessential example of system that
cannot be understood simply by iterative reduction of its components.
Emergent phenomena, mechanisms at different scales, etc.

Some have been explored in the ecological literature: (1) Some
applications of dynamic models in ecology assume long-run equilibrium.
(2) Stochasticity

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{2}
\tightlist
\item
  Ecological processes vary across more variables than the tools of
  analytic models are suited for. As the number of variables in an
  analytic model increases, so does the ability of the scientist to
  decern clear relationships between them, and so does overfitting
  potential. Curse of dimensionality--- Until the 20th century, no
  theory of the gravitational dynamics of more than 2 bodies.
  Understanding the gravitational dynamics of more than two planets with
  any reliability proved difficult. Using the same models (diffeqs), how
  could we adequately predict ecosystems?
\end{enumerate}

@@

The term \emph{ecological forecasting} implicitly creates an analogy
between predicting how ecosystems will change in the future by using the
term ``forecasting''---the most immediate analog being the success story
of weather forecasting via numerical weather prediction (NWP).

Although it is become almost hack to complain about the dang weather
forecast being wrong, over the least 50 years the (Bauer \emph{et al.}
2015).

The success of NWP, and the Earth observations that support it should
serve as a template for development of a system for monitoring Earth's
biodiversity. Much like ecology, NWP is faced with high-dimensional
systems that are governed by different mechanisms at different scales.

NWP has worked because it incorporates information about data and
meteorological processes collected at difference scales into models
that. Use of computational methods in NWP.

Much as one would not aim to forecast the weather in Quebec by applying
Navier-Stokes, forecasting ecological systems must

Transition to simulation as the solution: shift toward approach of
building models that \emph{generate} data.

(resolving the semantic ambuity of what differentiates ``mechanistic''
vs ``phenomological'' models is out of scope for now).

More broadly a reflection reflect ecology lagging behind the statistical
methods used in sciences that face similar challenges (many dimensions,
many mechanisms at different scales, each with stochasticity). Chaotic
dynamics emerge from simple analytic models, and . Whether ecosystems
actually exhibit chaotic behavior is a different question.

\textbf{P5}

But forecasting isn't the only difficult problem here.

Transition to theme of optimization given unknown information. A
forecast gives us a range of future values with uncertainty around them.
Further a convenient property that a forecasting model's uncertainty
goes up over time (if we assume the underlying process is Markov--this
is a strong assumption but oft true of the models we fit to temporal
data)

In face of uncertainty, decision making is an optimization problem. We
have some goal state for the future, and some estimate of what the state
of the world will be given a set of actions. Frame optimization problem
mathematically an introduce concept of solution-space and constraint.

Indeed Marx's most well known quote that ``philosophers have hitherto
only interpreted the world in various ways; the point is to change it.''
and a necessary step toward establishing a just and sustainable world.

@@

\textbf{P6 -- final intro para}

Three major components here: 1) Ecosystem monitoring, 2) Forecasting
using the products of that monitoring, and 3) Choosing the best possible
mitigation strategy.

This flow is outlined in the left panel of fig.~\ref{fig:thesis}

@@

The issue is that we don't really know what interactions are like now.
So not only do we need to predict with data that is spatially and
temporally sparse and likely to contain many interaction
``false-negatives'' (\textbf{Strydom2021?})

@@

\textbf{CH1 concept figure}

@@

The data for this chapter is derived from multiple souces and can be
split into three categories. (1) Field data from three different
locations acvross Colorado. All field sites have multiple plots across
an elevational gradient.

@@

This also goes on to includes testing some assumptions of the model with
empirical data fig.~\ref{fig:4}. which indicate our neutral model, if
anything, underestimates the probability of false-negatives due to
positive correlations in co-occurrence in two spatially replicated
networks (Thompson \& Townsend 2000; Hadfield \emph{et al.}
2014)---further I'm planning to add the field data from chapter one into
this anlysis once complete.

\begin{figure}
\centering
\includegraphics{./figures/positiveassociations.png}
\caption{f}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{./figures/software.png}
\caption{todo}
\end{figure}

@@

@@

@@

@@

@@

>>>>

Within the last several hundred years, human activity has induced rapid
changes in Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrain and oceans, and both agricultural and urban development has
rapidly reshaped the Earth's land cover. These the bulk of this change
has occurred within the last several hundred years, a geological
instant, inducing a sudden shift in conditions to Earth's climate and
biosphere. As a result, predicting how ecosystems will change in the
future, \emph{ecological forecasting}, and then using these forecasts to
make decisions to mitigate the negative consequences of this change on
ecosystems, their functioning, and the services they provide to humans
has emerged as an imperative for ecology and environmental science
(Dietze 2017). However, robust prediction of ecological processes is, to
say the least, quite difficult (Beckage \emph{et al.} 2011; Petchey
\emph{et al.} 2015). This difficultly is compounded by a few factors,
the first being that sampling ecosystems is not easy. Ecological data is
often biased, noisey, and sparse in both space and time. The current
paucity of ecological data has resulted in much interest in developing
global systems for \emph{ecosystem monitoring} (Makiola \emph{et al.}
2020), which would systematize the collection of biodiversity data in
manner that makes detecting and predicting change more possible than at
the moment (Urban \emph{et al.} 2021).

@@

The second major challenge in ecological forecasting is that the
underlying dynamics of most ecological processes are unknown and instead
must be inferred from this (sparse) data. Much of the history of
quantitatively modeling ecosystems have been done in the language of
dynamical systems, describing how the value of an observable state of
the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time, yielding
models in the form of differential equations in continuous-time
settings--\(\frac{dx}{dt} = f(x)\)-- or difference equations in
discrete-time settings--\(x_t = f(x_{t-1})\)--where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The initial success of these
forms of models can be traced back to the larger program of ontological
reductionism, which became the default approach to modeling in the
sciences after its early success in physics, which, by the time ecology
was becoming a quantitative science (sometime in the 20th century,
depending on who you ask), became the foundation for early quantitative
models in ecology.

@@

However, we run into many problems when aiming to apply this type of
model to empirical data in ecology. Ecosystems are perhaps the
quintessential example of system that cannot be understood by iterative
reduction of its components into constituent parts---ecological
phenomena are emergent are the product of different mechanisms operating
a different spatial, temporal, and organizational scales (Levin 1992).
Further, the form of this functional response in real systems is
effectively unknown, and some forms are inherently more ``forecastable''
than others (Beckage \emph{et al.} 2011; Chen \emph{et al.} 2019;
Pennekamp \emph{et al.} 2019). Further this analytical approach to
modeling explicitly ignores known realities: ecological dynamics not
deterministic, many analytic models in ecology assume long-run
equilibrium. Finally, perhaps the biggest challenge in using these
models to describe ecological processes is ecosystems vary across more
variables than the tools of analytic models are suited for. As the
number of variables in an analytic model increases, so does the ability
of the scientist to discern clear relationships between them given a
fixed amount of data, the so-called ``curse'' of dimensionality.

@@

But these problems are not solely unique to ecology. The term
\emph{ecological forecasting} implicitly creates an analogy with weather
forecasting. Although it has become a trite joke to complain about the
weather forecast being wrong, over the least 50 years the field of
numerical weather prediction (NWP) has dramatically improved out ability
to predict weather across the board (Bauer \emph{et al.} 2015). The
success of NWP, and the Earth observations systems that support it (Hill
\emph{et al.} 2004), should serve as a template for development of a
system for monitoring Earth's biodiversity. Much like ecology, NWP is
faced with high-dimensional systems that are governed by different
mechanisms at different scales. The success of NWP is that, rather than,
say, attempt to forecast the weather in Quebec by applying Navier-Stokes
to entire province, to instead use simulation models which describe
known mechanisms at different scales, and use the availability to
increasing computational power to directly simulate many batches of
dynamics which directly incorporate stochasticity and uncertainty in
parameter estimates via random number generation.

\textbf{P6}

But forecasting is only half the story. Marx's most well known quote
that ``philosophers have hitherto only interpreted the world in various
ways; the point is to change it.'' Indeed, once we have a forecast about
how an ecosystem will change in the future, what if this forecast
predicts a critical ecosystem service will deteriorate? We are still
left with the question, what do we in the time being to mitigate the
negative consequences a forecast predicts? In this framing, mitigating
the consequences of anthropogenic change on ecosystems becomes an
optimization problem: given a forecast of the probability. We have some
goal state for the future, and some estimate of what the state of the
world will be given a set of actions. Frame optimization problem
mathematically an introduce concept of solution-space and constraint.

@@

\textbf{P7}

This dissertation aims to formalize a framework for ecosystem monitoring
and forecasting (fig.~\ref{fig:thesis}, left), and each chapter address
some aspect of this pipeline to data from a monitoring network to
forecasts to mitigation strategy (fig.~\ref{fig:thesis}, right).

@@

The issue is that we don't really know what interactions are like now.
So not only do we need to predict with data that is spatially and
temporally sparse and likely to contain many interaction
``false-negatives'' (Strydom \emph{et al.} 2021).

@@

\begin{figure}
\centering
\includegraphics{./figures/ch1.png}
\caption{chapter one concept fig}
\end{figure}

@@

The data for this chapter is derived from multiple souces and can be
split into three categories. (1) Field data from three different
locations across Colorado. All field sites have multiple plots across an
elevational gradient.

@@

This also goes on to includes testing some assumptions of the model with
empirical data fig.~\ref{fig:posassoc}. which indicate our neutral
model, if anything, underestimates the probability of false-negatives
due to positive correlations in co-occurrence in two spatially
replicated networks (Thompson \& Townsend 2000; Hadfield \emph{et al.}
2014)---further I'm planning to add the field data from chapter one into
this anlysis once complete.

\begin{figure}
\hypertarget{fig:posassoc}{%
\centering
\includegraphics{./figures/positiveassociations.png}
\caption{f}\label{fig:posassoc}
}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{./figures/ch4.png}
\caption{todo}
\end{figure}

@@

\leavevmode\hypertarget{ref-Hill2004ArcEar}{}%
Hill, C., DeLuca, C., Balaji, Suarez, M. \& Da Silva, A. (2004). The
architecture of the Earth System Modeling Framework. \emph{Computing in
Science Engineering}, 6, 18--28.

@@

\leavevmode\hypertarget{ref-Levin1992ProPat}{}%
Levin, S.A. (1992). The Problem of Pattern and Scale in Ecology: The
Robert H. MacArthur Award Lecture. \emph{Ecology}, 73, 1943--1967.

@@

\leavevmode\hypertarget{ref-Pennekamp2019IntPre}{}%
Pennekamp, F., Iles, A.C., Garland, J., Brennan, G., Brose, U., Gaedke,
U., \emph{et al.} (2019). The intrinsic predictability of ecological
time series and its potential to guide forecasting. \emph{Ecological
Monographs}, 89, e01359.

@@

\leavevmode\hypertarget{ref-Strydom2021RoaPre}{}%
Strydom, T., Catchen, M.D., Banville, F., Caron, D., Dansereau, G.,
Desjardins-Proulx, P., \emph{et al.} (2021). \emph{A Roadmap Toward
Predicting Species Interaction Networks (Across Space and Time)}
(Preprint). EcoEvoRxiv.

@@

\leavevmode\hypertarget{ref-Urban2021CodLif}{}%
Urban, M.C., Travis, J.M.J., Zurell, D., Thompson, P.L., Synes, N.W.,
Scarpa, A., \emph{et al.} (2021). Coding for Life: Designing a Platform
for Projecting and Protecting Global Biodiversity. \emph{BioScience}.
